Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-06-27 NovelLBM.tex
Updates

The Lattice-Boltzmann method (LBM) has become a standard tool in numerical fluid mechanics \cite{Kruger_book}, and we therefore only mention the key choices here. Further details can be found in the appendix~\ref{app:LBM} or in literature.
We use a D3Q19 lattice in full three dimensions together with a single relaxation time collision operator.
Our implementation is based on FluidX3D which has been extensively used and validated in earlier publications \cite{lehmann2021, Lehmann.20229da}.
The core of the algorithm is implemented in OpenCL and thus capable of running on GPUs from different vendors.
Non-periodic boundaries are modeled using the half-way bounce back algorithm. To model fixed inflow/outflow, bounce back boundaries using a Dirichlet boundary condition as described by Krüger \textit{et al.} \cite{Kruger_book} are used.

>>>>

The Lattice-Boltzmann method (LBM) has become a standard tool in numerical fluid mechanics \cite{Kruger_book}, and we therefore only mention the key choices here. Further details can be found in the appendix~\ref{app:LBM} or in literature.
We use a D3Q19 lattice in full three dimensions together with a single relaxation time (SRT) collision operator.
\RR{Various tests with a multiple relaxation time (MRT) collision operator did not show significant differences.}
Our implementation is based on FluidX3D which has been extensively used and validated in earlier publications \cite{lehmann2021, Lehmann.20229da}.
The core of the algorithm is implemented in OpenCL and thus capable of running on GPUs from different vendors.
Non-periodic boundaries are modeled using the half-way bounce back algorithm. To model fixed inflow/outflow, bounce back boundaries using a Dirichlet boundary condition as described by Krüger \textit{et al.} \cite{Kruger_book} are used.

### 2024-06-27 responses/Response1.tex
Updates

The parameters $\eta_0$ and $\eta_\infty$ are the viscosities predicted by the model for zero and infinite shear rates, respectively.
We now mention this just before Eq. (7).\RK{Eigentlich beziehen sich die auf CY bzw. auf experimentelle Daten.}

@@

MRT is considered to improve numerical stability of time dependent or turbulent flow in high $Re$ cases \cite{dzanic2022, Kruger_book}, it is used for $Re = \num{100}$ \cite{chai2011} and above \cite{dhumieres2002}. As the problem discussed in the present work has very low $Re$, it would not be expected to improve stability here. Consequently, SRT is used by all authors we list in our literature overview except for Kuron et al. (TRT) \cite{kuron2021} and Gupta et al. (MRT) \cite{gupta2015}. Notably, while Gupta et al. use MRT their method achieves the lowest $max\left(R_\eta\right)$ in our overview. We do not want to imply, that MRT might be counterproductive here (Gupta et al. did not compare to SRT), but it is also clearly no remedy. That being said, given, that we have MRT implemented anyway, we of course tried it before coming up with the present algorithm and were not able to observe any improvement.

>>>>

The parameters $\eta_0$ and $\eta_\infty$ are the viscosities for zero and infinite shear rates, respectively.
We now mention this just before Eq. (7).

@@

MRT is considered to improve numerical stability of time dependent or turbulent flow in high $Re$ cases \cite{dzanic2022, Kruger_book}, it is used for $Re = \num{100}$ \cite{chai2011} and above \cite{dhumieres2002}. 
As the problem discussed in the present work has very low $Re$, it would not be expected to improve stability here. 
Consequently, SRT is used by all authors we list in our literature overview except for Kuron et al. (TRT) \cite{kuron2021} and Gupta et al. (MRT) \cite{gupta2015}. 
Notably, while Gupta et al. use MRT their method achieves the lowest $max\left(R_\eta\right)$ in our overview, which leads us to believe that MRT does not lead to stability improvement for low Reynolds viscoelastic situations such as ours.

To confirm the above hints from the literature, we have done ourselves some tests with MRT (which is implemented in our code) but could not find any stability improvement. 
We now mention this in the revised manuscript in section 4.1.

### 2024-06-27 responses/Response2.tex
Updates

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{Submission ID \SG{ to fill }}
\rfoot{\thepage}
\setlength\parindent{0pt}

\newcommand{\SG}[1]{{\color{red}SG:~#1}}
\newcommand{\SJM}[1]{{\color{orange}SM:~#1}}
\newcommand{\SW}[1]{{\color{purple}SW:~#1}}

@@

We agree with the referee that the number of $2 \cdot 10^9$ time steps may appear rather large in the first place.
However, LBM schemes can be implemented in an extremely efficient manner, especially using GPUs (note that our code, both the LBM and the finite-volume part, is entirely GPU-based).
The longest simulation presented in the manuscript which is for a fully 3D cylindrical flow took 1 day and 8 hours on a five-year old gaming GPU.

That said, the referee is right that highly viscous fluids such as, in particular, the alginate solution considered in this work, are a challenge to all LBM schemes.
We hypothesize that this is one of the reasons why such highly viscous complex fluids have been simulated with LBM before.
A possible remedy which may be considered in future work is the so-called Reynolds scaling.
Herein, the Reynolds number of the flow is artificially increased which decreases the LBM runtime without modifying the physical flow field as long as Re remains (much) smaller than 1.
We now mention this possibility above Eq. (13) in the manuscript.

@@

Aside from figures 3 and 5, which are planar geometries, all our systems are fully 3D.
This includes the cylindrical flow in figure 4. 
Even more important, it includes the RT-DC geometry shown in figure 7 and the fully three-dimensional cell in figure 8.
To bring out more clearly the latter aspect, we have added to figure 8 an illustration of our simulated system \SG{ denote current figure as 8b, and add a standard 3D Paraview picture as 8a. }

>>>>

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{Submission ID FLD-24-0080}
\rfoot{\thepage}
\setlength\parindent{0pt}

\newcommand{\SG}[1]{{\color{red}SG:~#1}}
\newcommand{\SJM}[1]{{\color{orange}SM:~#1}}
\newcommand{\SW}[1]{{\color{purple}SW:~#1}}
\newcommand{\RK}[1]{{\color{cyan}RK: #1}}

@@

We agree with the referee that the number of $2 \cdot 10^9$ time steps may appear rather large in the first place.
However, LBM schemes can be implemented in an extremely efficient manner, especially using GPUs (note that our code, both the LBM and the finite-volume part, is entirely GPU-based).
The longest simulation presented in the manuscript which is for a fully 3D cylindrical flow took 1 day and 8 hours on a five-year old gaming GPU.
%\RK{Wohlgemerkt war das der 3D Poisueille fürs Alginat (was das eigentliche Problem ist, das RT-DC mesh ist viel größer, haben wir aber mit MC gerechnet, hat 40 Minuten gedauert). Für die MC hat der 3D Poiseuille nicht ganz 5 Minuten gedauert.}

That said, the referee is right that highly viscous fluids such as, in particular, the alginate solution considered in this work, are a challenge to all LBM schemes.
%We hypothesize that this is one of the reasons why such highly viscous complex fluids have not been simulated with LBM before.\RK{Fehlt da ein not? Ist dann inhaltlich aber auch falsch, weil Re scaling. Es wurde nicht simuliert weil nicht stabil}
A possible remedy which may be considered in future work is the so-called Reynolds scaling.
Herein, the Reynolds number of the flow is artificially increased which decreases the LBM runtime without modifying the physical flow field as long as Re remains (much) smaller than 1.
We now mention this possibility above Eq. (13) in the manuscript.

@@

Aside from figures 3 and 5, which are planar geometries, all our systems are fully 3D.
This includes the cylindrical flow in figure 4. 
Even more important, it includes the RT-DC geometry shown in figure 7 and the fully three-dimensional cell in figure 8.
To bring out more clearly the latter aspect, we have added to figure 8 an illustration of our simulated system.
